### 2024-05-09 paper/tex/01-introduction.tex
amend

\paragraph{Declarative Type System Specification (\cref{sec:declarative-system,sec:proofs})} 
    We devise \fexists---a second-order polymorphic type system that supports first-class
    existential and universal quantifiers, both of which can be instantiated
    impredicatively. To evade the undecidability results surrounding subtyping
    for \systemf, we base the system on \CBPV \cite{levy2006:cbpv}: we
    use \emph{polarization} of types and terms to formulate the restrictions
    that make typing decidable.

@@

\paragraph{Theoretical results: correctness of the algorithm (\cref{sec:proofs})}
  We prove that our algorithm is terminating and both sound and complete with
  respect to the declarative specification. This proof of soundness and completeness
  of the algorithm required us to introduce an intermediate system in-between
  the declarative and the algorithmic systems, and furthermore, the
  soundness and completeness proofs are mutually recursive with each other.

@@

% \ilyam{Is the following needed since we have the contributions? I just wanted to introduce the appendix somewhere...}
\paragraph{\textbf{Structure of this Paper}}
In \cref{sec:overview}, we discuss the language of the \fexists system, providing
examples, outlining limitations, and summarizing the main ideas behind the
algorithm. \Cref{sec:declarative-system} presents a formal definition of the
declarative specification and its key properties. In \cref{sec:algorithm}, the
type inference algorithm is introduced, detailing a hierarchy of subroutines;
upon perusing this section, one could \emph{implement} the type inference
algorithm. \Cref{sec:proofs} overviews the algorithm's correctness proof,
explaining its structure and how the declarative system's features make it
decidable. \Cref{sec:extensions} covers potential \fexists extensions and future
work directions. Finally, \cref{appendix} contains comprehensive formal
definitions and proofs of lemmas and theorems used in the paper; in instances
where the full definitions might not be of immediate interest, we direct the
reader to the appendix.

>>>>

\paragraph{Declarative Type System Specification (\cref{sec:overview,sec:declarative-system})} 
    We devise \fexists---a second-order polymorphic type system that supports first-class
    existential and universal quantifiers, both of which can be instantiated
    impredicatively. To evade the undecidability results surrounding subtyping
    for \systemf, we base the system on \CBPV \cite{levy2006:cbpv}: we
    use \emph{polarization} of types and terms to formulate the restrictions
    that make typing decidable.

@@

\paragraph{Theoretical results: correctness of the algorithm (\cref{sec:proofs})}
  We prove that our algorithm is terminating and both sound and complete with
  respect to the declarative specification. This proof of soundness and completeness
  of the algorithm required us to introduce an intermediate system in-between
  the declarative and the algorithmic systems, and furthermore, the
  soundness and completeness proofs are mutually recursive with each other.
  The central theorems and the key proof ideas are formulated in \cref{sec:proofs}, 
  while the full definitions and formal proofs are deferred to the appendix (\cref{appendix}).

@@

% \ilyam{Is the following needed since we have the contributions? I just wanted to introduce the appendix somewhere...}
% \paragraph{\textbf{Structure of this Paper}}
% In \cref{sec:overview}, we discuss the language of the \fexists system, providing
% examples, outlining limitations, and summarizing the main ideas behind the
% algorithm. \Cref{sec:declarative-system} presents a formal definition of the
% declarative specification and its key properties. In \cref{sec:algorithm}, the
% type inference algorithm is introduced, detailing a hierarchy of subroutines;
% upon perusing this section, one could \emph{implement} the type inference
% algorithm. \Cref{sec:proofs} overviews the algorithm's correctness proof,
% explaining its structure and how the declarative system's features make it
% decidable. \Cref{sec:extensions} covers potential \fexists extensions and future
% work directions. Finally, \cref{appendix} contains comprehensive formal
% definitions and proofs of lemmas and theorems used in the paper; in instances
% where the full definitions might not be of immediate interest, we direct the
% reader to the appendix.

### 2024-05-09 paper/tex/02-overview.tex
amend

To check whether one type is a subtype of another, the algorithm introduces
\emph{algorithmic} type variables ($[[α̂±]], [[β̂±]], \dots$)--- the
`placeholders' representing the polymorphic variables whose instantiation is
postponed. This way, the problem of $[[∀α⁺.↑α⁺ ≤ ↑Int]]$ becomes the problem
of finding $[[α̂⁺]]$ such that $[[↑α̂⁺ ≤ ↑Int]]$.

>>>>

To check whether one type is a subtype of another, the algorithm introduces
\emph{algorithmic} type variables ($[[α̂±]], [[β̂±]], \dots$)---the
`placeholders' representing the polymorphic variables whose instantiation is
postponed. This way, the problem of $[[∀α⁺.↑α⁺ ≤ ↑Int]]$ becomes the problem
of finding $[[α̂⁺]]$ such that $[[↑α̂⁺ ≤ ↑Int]]$.

### 2024-05-09 paper/tex/03-declarative-system.tex
amend

The normalization algorithm is shown in \cref{fig:type-nf}.
The steps (i-ii) are implemented by the \emph{ordering} function
`$[[ord varset in iN]]$' and `$[[ord varset in iP]]$'.
For a set of variables $[[varset]]$, and a type, it 
returns a list of variables from $[[varset]]$ that occur in the type in 
the \emph{order of their first occurrence}.
For brevity, we omit the formal definition of ordering referring the reader to the appendix (\cref{sec:ordering}).

@@

The type polarization (\cref{fig:polarization}) is more involved, as there are multiple
ways to polarize a type: for instance, a variable $[[α]]$ can be polarized to $[[α⁺]]$ or $[[α⁻]]$.
The choice of polarization affects the execution strategy of the program.
\ilyam{How?}
We chose the positive polarization: every \systemf type is translated to a \emph{positive} type of \fexists.
This choice makes it smoother to lift the translation to the term level, which will be disused next.

@@

For brevity, we omit the formal definition of the term-level elaboration,
referring the reader to the appendix (\cref{sec:term-translation}). We define
the elaboration in such a way that it preserves the soundness invariants given
in \cref{lemma:translation-soundness}.

\begin{table}[h]
\centering
\begin{tabular}{rll}
\hline
\multicolumn{2}{c}{\textbf{Elaboration Judgment}} & \textbf{The Soundness Property} \\
\hline
Negative Subtyping & $[[Γ ⊢ iN ≤ iM ⤳ ft]]$ & $[[|Γ| ; · ⊢ ft : |iN| → |iM| ]]$ \\
Positive Subtyping & $[[Γ ⊢ iP ≥ iQ ⤳ ft]]$ & $[[|Γ| ; · ⊢ ft : |iQ| → |iP| ]]$ \\
Negative Typing & $[[Γ; Φ ⊢ c : iN ⤳ ft]]$ & $[[|Γ| ; |Φ| ⊢ ft : |iN|]]$ \\
Positive Typing & $[[Γ; Φ ⊢ v : iP ⤳ ft]]$ & $[[|Γ| ; |Φ| ⊢ ft : |iP|]]$ \\
Application Typing & $[[Γ ; Φ ⊢ iN ● args ⇒> iM ⤳ fe; fargs]]$ & $[[|Γ| ; |Φ|, x:|iN| ⊢ fe (x ● ft *) : |iM|]]$ \\
\hline
\systemf Typing & $[[fΓ ; fΦ ⊢ ft : fT ⤳ c]]$ & $[[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑+|fT|+]]$ \\
\hline
\end{tabular}
\caption{Soundness of the Elaboration:
  the elaboration judgment on the left implies the typing judgment on the right}
\end{table}

>>>>

The normalization algorithm is shown in \cref{fig:type-nf}. The steps (i-ii) are
implemented by the \emph{ordering} function `$[[ord varset in iN]]$' and `$[[ord
varset in iP]]$'. For a set of variables $[[varset]]$, and a type, it returns a
list of variables from $[[varset]]$ that occur in the type in the \emph{order of
their first occurrence}. For brevity, we omit the formal definition of ordering
referring the reader to the appendix (\cref{appendix}).

@@

The type polarization (\cref{fig:polarization}) is more involved, as there are
multiple ways to polarize a type: for instance, a variable $[[α]]$ can be
polarized to $[[α⁺]]$ or $[[α⁻]]$. The choice of polarization affects the
execution strategy of the program. 
In particular, a functional type can be represented in either positive (thunked, call-by-name) way:
$[[↓(iP → ↑iQ)]]$ or negative (call-by-value): $[[↓iN → iM]]$.
We chose the positive polarization: every \systemf type is translated to a
\emph{positive} type of \fexists. This choice makes it smoother to lift the
translation to the term level, which will be discussed next.

@@

For brevity, we omit the formal definition of the term-level elaboration,
referring the reader to the appendix (\cref{sec:term-translation}). We define
the elaboration in such a way that it preserves the soundness invariants given
in \cref{tab:elaboration-soundness}.

\begin{table}[h]
\centering
\begin{tabular}{rll}
\hline
\multicolumn{2}{c}{\textbf{Elaboration Judgment}} & \textbf{The Soundness Property} \\
\hline
Negative Subtyping & $[[Γ ⊢ iN ≤ iM ⤳ ft]]$ & $[[|Γ| ; · ⊢ ft : |iN| → |iM| ]]$ \\
Positive Subtyping & $[[Γ ⊢ iP ≥ iQ ⤳ ft]]$ & $[[|Γ| ; · ⊢ ft : |iQ| → |iP| ]]$ \\
Negative Typing & $[[Γ; Φ ⊢ c : iN ⤳ ft]]$ & $[[|Γ| ; |Φ| ⊢ ft : |iN|]]$ \\
Positive Typing & $[[Γ; Φ ⊢ v : iP ⤳ ft]]$ & $[[|Γ| ; |Φ| ⊢ ft : |iP|]]$ \\
Application Typing & $[[Γ ; Φ ⊢ iN ● args ⇒> iM ⤳ fe; fargs]]$ & $[[|Γ| ; |Φ|, x:|iN| ⊢ fe (x ● ft *) : |iM|]]$ \\
\hline
\systemf Typing & $[[fΓ ; fΦ ⊢ ft : fT ⤳ c]]$ & $[[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑+|fT|+]]$ \\
\hline
\end{tabular}
\caption{Soundness of the Elaboration:
  the elaboration judgment on the left implies the typing judgment on the right}
  \label{tab:elaboration-soundness}
\end{table}

### 2024-05-09 paper/tex/06-extensions.tex
amend

After the weakening of subtyping invariants, 
it is possible to smoothly extend the \emph{elementary}
version of the system with bounded quantification.
In particular, we can add upper bounds to polymorphic $[[∀]]$-quantifiers:
$[[∀(pa ≤* iP). iN]]$.

>>>>

After the weakening of subtyping invariants, 
it is possible to smoothly extend the \emph{elementary}
version of the system 
(the second solution discussed in \cref{sec:explicit-type-application})
 with bounded quantification.
In particular, we can add upper bounds to polymorphic $[[∀]]$-quantifiers:
$[[∀(pa ≤* iP). iN]]$.
